### 2018-10-10 algorithm.tex
fix #9: average nu across countries
free trade formulas

\documentclass[12pt]{article}
\usepackage{amsmath}
\begin{document}
Solving a simple Armington-endowment model.

Let $Q_n = A_NL_n$, with $L_n$ completely inelastic. Demand is CES:
\[
C_n^{1-1/\theta} = \sum_m Q_m^{1-1/\theta}.
\]
Trade costs are $\tau_{ij}$ and we denote $\kappa_{ij} = \tau_{ij}^{-\theta}$.

This model has the following equilibrium equations.
\[
d_{nm} = 
	\frac
	{(\tau_{nm}w_m)^{1-\theta}}
	{\sum_k (\tau_{nk} w_k)^{1-\theta}}
\]
\[
w_m \cdot L_m = \sum_n d_{nm} w_n \cdot L_n
\]
where $w_m$ is the unit cost of labor $L_m$.
\[
w_m L_m = w_m^{1-\theta}
\sum_n 	\frac
	{\tau_{nm}^{1-\theta}}
	{\sum_k (\tau_{nk} w_k)^{1-\theta}}
 w_n L_n
\]
Let $\Phi_n = \sum_k (\tau_{nk} w_k)^{1-\theta}$ 
\[
w_m^\theta L_m = 
\sum_n 	
	{\tau_{nm}^{1-\theta}}
 w_n L_n/\Phi_n.
\]
This leads to a better iteration: Pick $w_n$ and calculate corresponding $\Phi_n$. Calculate new $w_n$ from LHS.

\begin{enumerate}
	\item Pick $\alpha_m^{(0)} = w_mL_m/\Phi_m$ from some previous equilibrium (or data).
	\item Solve for
	\[
	w_m^{(k+1)} = L_m^{-1/\theta} 
		\left[
		\sum_n \alpha_n^{(k)} \tau_{nm}^{1-\theta}
		\right]^{1/\theta}
	\]
	\item Set 
	\[
	\alpha_m^{(k+1)} = w_m^{(k)} L_m/\Phi_m(w^{(k)})
	\]
	and repeat.
\end{enumerate}
This should converge fast, in 3-4 steps, because (i) we start from a close guess, (ii) $\theta$ is large so that $1/\theta$ is small, resulting in dampening, (iii) $\alpha$ has limited impact on wages, only due to destination-country variation in $\tau$, dampening feedback.

\section{Full model}
Introduce a new variable for the factory-gate price of intermediate goods,
\begin{equation}\label{eq:varrho}
	\varrho_{njt} = \xi B_j T_{nj}^{-1/\theta} A_{njt}^{-1} w_{njt}^{\beta_j} \prod_{k=1}^J P_{nkt}^{\gamma_{kj}}.
\end{equation}
With this variable, we can write the price of the final good as
\begin{equation}\label{eq:price}
	P_{mjt} = \left[
		\sum_{n=1}^N (\varrho_{njt}/\kappa_{mnjt})^{-\theta}
		\right]^{-1/\theta}.
\end{equation}
This is a standard CES price index. We can write import shares as
\begin{equation}\label{eq:import_share}
	d_{mnjt} = \left[
	\frac
	{\varrho_{njt}/\kappa_{mnjt}}
	{P_{mjt}}
	\right]^{-\theta}.
\end{equation}
Wages can be expressed by inverting \eqref{eq:varrho},
\begin{equation}\label{eq:wage}
	w_{njt} = (\xi B_j)^{-1/\beta_j}
	\varrho_{njt}^{1/\beta_j}  
	T_{nj}^{1/(\beta_j\theta)}A_{njt}^{1/\beta_j} 
	\prod_{k=1}^J P_{nkt}^{-\gamma_{kj}/\beta_j}.
\end{equation}

@@

Given $d_{njt}$ and the precomputed $E_{wjt}$, we can calculate wages in all sectors. Prices can be calculated from
\[
\ln\mathbf P_{t} = 
(\mathbf I-\Gamma)^{-1}[\text{diag}(\beta+1/\theta) \ln\mathbf d_{0t}  
+\ln\xi
	+ \ln\mathbf B
	- \ln\mathbf A_{0t}
+
\text{diag}(\beta)(\ln\mathbf E_{wt}-\ln\mathbf L_{0t})]
 .  
\]

\section{Expectations and outer loop}
An equilibrium condition states
\begin{equation}
	\frac 
		{L_{njt}}
		{L_{nt}}
	=
	E_{t-1}\frac 
		{w_{njt}L_{njt}}
		{w_{nt}L_{nt}}.
\end{equation}
We calculate the expected wage share as follows. Given past productivities $A_{nj,t-1}$, draw $S$ realizations for current productivities, $A_{njs}$. Because we are looking for a share bounded between 0 and 1, we don't need a sophisticated sampling to calculate the numerical integral.

@@

\subsection{Time series}
Given the autoregressive nature of productivity, the vector of $A_{jnt}$s is a state variable. The problem is drastically simplified by the fact that no decision variable affects the state. The representative agent solves
% FIXME: we have already used varrho
\[
\max_{\{L_{njt}, L_{njt}^*\}}
E
\sum_{t=0}^{\infty} 
	\delta^{t} 
	\left[
	\ln \frac 
		{\sum_j w_{njt}(\mathbf A_{t}, \mathbf L_{t})L_{njt}}
		{P_{nt}(\mathbf A_{t}, \mathbf L_{t})}
	-\frac \varrho 2
		\sum_{j=1}^J
		\left(
			\frac {L_{njt}} {L_{nt}}
			- 
			\frac {L_{njt}^*} {L_{nt}}
		\right)^2
	\right]
\]
subject to $\sum_j L_{njt}=\sum_j L_{njt}^*=L_{nt}$ and $w$ and $P$ being determined in equilibrium. The starred labor allocation is determined before shock realizations are known.

@@

The night labor share is a fixed point of
\[
E_{A}\frac 
	{\tilde w_{njt}(\mathbf A_{t}, \mathbf L_{t}^*)}
	{\tilde w_{nt}(\mathbf A_{t}, \mathbf L_{t}^*)}
= \lambda L_{nt}.
\]
\subsection{Calibrating the CES model}
To calibrate the CES demand shifters $\nu$, we need data on how final expenditure is split across sectors in each country. We will recover this from the split of revenue, subtracting intermediate expenditure.

@@

Computing the price index for US, 1972,
\begin{equation}
	\left[\sum_{j=1}^J \nu_{US,jt} 1^{1-\sigma}\right]^{1/(1-\sigma)} = 
	\left[\sum_{j=1}^J \nu_{US,jt}\right]^{1/(1-\sigma)}
 =1,
\end{equation}
so we are measuring all nominal quantities in 1972 US dollars.

>>>>

\documentclass[12pt]{article}
\usepackage{amsmath}
\begin{document}
\section{Model solution}
Introduce a new variable for the factory-gate price of intermediate goods,
\begin{equation}\label{eq:varrho}
	\varrho_{njt} = \xi B_j T_{nj}^{-1/\theta} A_{njt}^{-1} w_{njt}^{\beta_j} \prod_{k=1}^J P_{nkt}^{\gamma_{kj}}.
\end{equation}
With this variable, we can write the price of the final good as
\begin{equation}\label{eq:price}
	P_{mjt} = \left[
		\sum_{n=1}^N (\varrho_{njt}/\kappa_{mnjt})^{-\theta}
		\right]^{-1/\theta}.
\end{equation}
This is a standard CES price index. We can write import shares as
\begin{equation}\label{eq:import_share}
	d_{mnjt} = \left[
	\frac
	{\varrho_{njt}/\kappa_{mnjt}}
	{P_{mjt}}
	\right]^{-\theta}.
\end{equation}
Wages can be expressed by inverting \eqref{eq:varrho},
\begin{equation}\label{eq:wage}
	w_{njt} = (\xi B_j)^{-1/\beta_j}
	\varrho_{njt}^{1/\beta_j}  
	T_{nj}^{1/(\beta_j\theta)}A_{njt}^{1/\beta_j} 
	\prod_{k=1}^J P_{nkt}^{-\gamma_{kj}/\beta_j}.
\end{equation}

@@

Given $d_{njt}$ and the precomputed $E_{wjt}$, we can calculate wages in all sectors. 
\[
-\frac1\theta \ln\mathbf d_{nt} = 
\ln\xi
	+ \ln\mathbf B
	- \ln\mathbf A_{nt}
+\text{diag}(\beta)(\ln\mathbf w_{nt}-\ln\mathbf P_{t})
+ \mathbf C \ln\mathbf P_{t}  
\]
\[
[\text{diag}(\beta)
- \mathbf C] \ln\mathbf P_{t} =  
\frac1\theta \ln\mathbf d_{nt} +
\ln\xi
	+ \ln\mathbf B
	- \ln\mathbf A_{nt}
+\text{diag}(\beta)\ln\mathbf w_{nt}
\]
Prices can be calculated from
\[
\ln\mathbf P_{t} =  
(\mathbf I - \Gamma)^{1}
[\frac1\theta \ln\mathbf d_{nt} +
\ln\xi
	+ \ln\mathbf B
	- \ln\mathbf A_{nt}
+\text{diag}(\beta)\ln\mathbf w_{nt}]
\]
\textbf{Note: This gamma matrix is the transpose of gammajk in the code.}

\subsection{Expectations and outer loop}
An equilibrium condition states
\begin{equation}
	\frac 
		{L_{njt}}
		{L_{nt}}
	=
	E_{t-1}\frac 
		{w_{njt}L_{njt}}
		{w_{nt}L_{nt}}.
\end{equation}
We calculate the expected wage share as follows. Given past productivities $A_{nj,t-1}$, draw $S$ realizations for current productivities, $A_{njs}$. Because we are looking for a share bounded between 0 and 1, we don't need a sophisticated sampling to calculate the numerical integral.

@@

\subsubsection{Time series}
Given the autoregressive nature of productivity, the vector of $A_{jnt}$s is a state variable. The problem is drastically simplified by the fact that no decision variable affects the state. The representative agent solves
% FIXME: we have already used varrho
\[
\max_{\{L_{njt}, L_{njt}^*\}}
E
\sum_{t=0}^{\infty} 
	\delta^{t} 
	\left[
	\ln \frac 
		{\sum_j w_{njt}(\mathbf A_{t}, \mathbf L_{t})L_{njt}}
		{P_{nt}(\mathbf A_{t}, \mathbf L_{t})}
	-\frac \varrho 2
		\sum_{j=1}^J
		\left(
			\frac {L_{njt}} {L_{nt}}
			- 
			\frac {L_{njt}^*} {L_{nt}}
		\right)^2
	\right]
\]
subject to $\sum_j L_{njt}=\sum_j L_{njt}^*=L_{nt}$ and $w$ and $P$ being determined in equilibrium. The starred labor allocation is determined before shock realizations are known.

@@

The night labor share is a fixed point of
\[
E_{A}\frac 
	{\tilde w_{njt}(\mathbf A_{t}, \mathbf L_{t}^*)}
	{\tilde w_{nt}(\mathbf A_{t}, \mathbf L_{t}^*)}
= \lambda L_{nt}.
\]

\subsection{Flexible labor equilibrium}
To get a starting point or direction of iteration, it is useful to calculate the flexible labor equilibrium, where wages are equal across sectors.

\section{Calibration}
\subsection{Calibrating the CES model}
To calibrate the CES demand shifters $\nu$, we need data on how final expenditure is split across sectors in each country. We will recover this from the split of revenue, subtracting intermediate expenditure.

@@

Computing the price index for US, 1972,
\begin{equation}
	\left[\sum_{j=1}^J \nu_{US,jt} 1^{1-\sigma}\right]^{1/(1-\sigma)} = 
	\left[\sum_{j=1}^J \nu_{US,jt}\right]^{1/(1-\sigma)}
 =1,
\end{equation}
so we are measuring all nominal quantities in 1972 US dollars. We then average across countries to get $\nu_{jt}$.
